---
title: "Comparing kNN and DANN"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Comparing kNN and DANN}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Introduction

Discriminant Adaptive Nearest Neighbor (DANN) classification is an enhancement of standard k-nearest neighbors (kNN) that adapts the distance metric based on local class structure. This vignette demonstrates when and why DANN outperforms standard kNN.

## The Key Idea

Standard kNN uses Euclidean distance, treating all directions equally. This works well when classes are roughly spherical, but struggles when:

1. **Classes have different covariance structures** - one class might be elongated in one direction, another in a different direction
2. **Irrelevant features exist** - noise dimensions dilute the signal from informative features

DANN addresses this by computing a local linear discriminant analysis around each query point, then modifying the distance metric to:

- Stretch neighborhoods along class boundaries (where discrimination is difficult)
- Shrink neighborhoods perpendicular to boundaries (where discrimination is easy)

## Example 1: Different Covariance Structures

Consider two classes where each is elongated, but in different directions:

```{r covariance-example}
library(dann)
set.seed(42)

n <- 150

# Class 1: elongated horizontally (high variance in x, low in y)
class1_x <- rnorm(n, mean = 0, sd = 2.0)
class1_y <- rnorm(n, mean = 0, sd = 0.4)

# Class 2: elongated vertically (low variance in x, high in y)
class2_x <- rnorm(n, mean = 0, sd = 0.4)
class2_y <- rnorm(n, mean = 0, sd = 2.0)

# Combine
x <- rbind(
  cbind(class1_x, class1_y),
  cbind(class2_x, class2_y)
)
y <- c(rep(1L, n), rep(2L, n))

# Shuffle
idx <- sample(2 * n)
x <- x[idx, ]
y <- y[idx]

# Split train/test
train_idx <- 1:240
test_idx <- 241:300

x_train <- x[train_idx, ]
x_test <- x[test_idx, ]
y_train <- y[train_idx]
y_test <- y[test_idx]

# Standardize
std <- stand(x_train, x_test)

# Compare methods
knn_pred <- knn(std$x, std$xx, y_train, k = 5)
dann_pred <- dann(std$x, std$xx, y_train, k = 5, kmetric = 40, fullw = TRUE)

knn_acc <- mean(knn_pred == y_test)
dann_acc <- mean(dann_pred == y_test)

cat("5-NN accuracy: ", round(knn_acc * 100, 1), "%\n", sep = "")
cat("DANN accuracy: ", round(dann_acc * 100, 1), "%\n", sep = "")
```

Both methods can handle this case reasonably well because the classes are geometrically distinguishable, but DANN's adaptive metric helps it find the optimal local decision boundary.

## Example 2: Irrelevant Features

A more challenging scenario is when only some features contain class information:

```{r noise-example}
set.seed(123)
n <- 150

# 2 informative features: classes separated
x_informative <- rbind(
  matrix(rnorm(n * 2, mean = 0, sd = 0.5), n, 2),
  matrix(rnorm(n * 2, mean = 2, sd = 0.5), n, 2)
)

# 8 noise features: pure random, no class information
x_noise <- matrix(rnorm(2 * n * 8, sd = 1), 2 * n, 8)

x <- cbind(x_informative, x_noise)
y <- c(rep(1L, n), rep(2L, n))

# Shuffle and split
idx <- sample(2 * n)
x <- x[idx, ]
y <- y[idx]

train_idx <- 1:240
test_idx <- 241:300

# Standardize
std <- stand(x[train_idx, ], x[test_idx, ])

# Compare
knn_pred <- knn(std$x, std$xx, y[train_idx], k = 5)
dann_pred <- dann(std$x, std$xx, y[train_idx], k = 5, kmetric = 40, fullw = TRUE)

knn_acc <- mean(knn_pred == y[test_idx])
dann_acc <- mean(dann_pred == y[test_idx])

cat("With 8 noise features:\n")
cat("5-NN accuracy: ", round(knn_acc * 100, 1), "%\n", sep = "")
cat("DANN accuracy: ", round(dann_acc * 100, 1), "%\n", sep = "")
```

DANN's local discriminant analysis helps it down-weight the noise dimensions.

## Real Data: Glass Classification

The package includes the glass identification dataset:

```{r glass-example}
data(dannt)

# Standardize
std <- stand(glass.train$x, glass.test$x)

# Methods
knn_pred <- knn(std$x, std$xx, glass.train$y, k = 5)
dann_pred <- dann(std$x, std$xx, glass.train$y, k = 5, kmetric = 50)

knn_errors <- sum(knn_pred != glass.test$y)
dann_errors <- sum(dann_pred != glass.test$y)

n_test <- length(glass.test$y)

cat("Glass dataset (", n_test, " test samples, 6 classes):\n", sep = "")
cat("5-NN errors: ", knn_errors, " (", round(100 * knn_errors / n_test, 1), "%)\n", sep = "")
cat("DANN errors: ", dann_errors, " (", round(100 * dann_errors / n_test, 1), "%)\n", sep = "")
```

## Parameters

### k and kmetric

- `k`: Number of neighbors for the final classification vote (default: 5)
- `kmetric`: Number of neighbors used to estimate the local covariance/metric (default: 50 or 20% of n)

A larger `kmetric` gives a smoother metric estimate but may miss local structure.

### epsilon

The softening parameter controls how much the metric adapts:

```{r epsilon-example}
data(dannt)
std <- stand(glass.train$x, glass.test$x)

# Try different epsilon values
epsilons <- c(0.1, 1, 10)
results <- dann(std$x, std$xx, glass.train$y, k = 5, kmetric = 50, epsilon = epsilons)

for (i in seq_along(epsilons)) {
  errors <- sum(results[, i] != glass.test$y)
  cat("epsilon =", epsilons[i], "-> errors:", errors, "\n")
}
```

- Small epsilon: more adaptation (metric changes more based on local structure)
- Large epsilon: less adaptation (closer to standard Euclidean metric)

### fullw and scalar

- `fullw = FALSE` (default): Use diagonal covariance (faster, often sufficient)
- `fullw = TRUE`: Use full covariance matrix (captures correlations between features)
- `scalar = TRUE`: Use scalar (isotropic) covariance (simplest, like weighted Euclidean)

## References

Hastie, T., & Tibshirani, R. (1996). Discriminant Adaptive Nearest Neighbor Classification. *IEEE Transactions on Pattern Analysis and Machine Intelligence*, 18(6), 607-616.
